Computer-Assisted Analysis of Emendations of G\"odel's Ontological Proof

### 2014-10-20 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
Reduce redundancy and boldface logic names

\usepackage{url}
\usepackage{commands}
\usepackage{xcolor}
\usepackage{comment}

@@

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail what is called the \emph{modal collapse}
\cite{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument and stimulated attempts to remedy the problem. Hájek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996} proposed the use
of cautious instead of full comprehension principles, and Fitting \cite{fitting02:_types_tableaus_god} suggested that greater care is necessary in the semantics of higher-order quantifiers in the presence of modalities. Others, such as
Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, proposed slight emendations of
G\"odel's axioms and definitions. They require neither comprehension restrictions nor more complex semantics. Therefore, they are technically simpler to analyze with computer support. We have formalized them using the proof
assistant Isabelle/HOL \cite{Isabelle} together with the automated higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof}, and Nitpick \cite{Nitpick}.  
Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced in previous
work \cite{C40,J30,J23}. We explored the effect of different domain conditions
 on the provability of lemmas, theorems and even axioms. This was motivated by a controversy between Hájek and Anderson regarding the redundancy of some axioms in Anderson's emendation. In \emph{constant domain semantics}, the individual domains are the same in all possible worlds. In \emph{varying domain semantics}, the domains may vary from world to world. This variation is technically encoded with the help of an existence relation expressing which individuals actually exist in each world. Quantifiers are then uniformly formalized as \emph{actualist quantifiers} (i.e. guarded by the existence relation). Our main results are summarized in the following sections.

@@

\BWP{Bruno: maybe we should be more explicit about these
comments}

\paragraph{Constant domain semantics}
%\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_constant_domain.thy}} 
     G\"odel's theorems T1, C
  and T3' can be quickly automated (in logics K and KB, respectively),
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B and the latter one already in K). Moreover, a trivial
  countermodel (consisting of two worlds and two individuals) to MC
  generated by Nitpick (for all mentioned logics); this model also
  shows the consistency of the axioms.

\paragraph{Varying domain semantics}
%\footnote{Cf. Isabelle/HOL source file \url{Anderson/Anderson_varying_domain.thy}} 
: In this setting we obtain the
  same results as above.

@@

\paragraph{Constant domain semantics}
%\footnote{Cf. Isabelle/HOL source files \url{Bjordal/Bjordal_A/B/C_constant_domain.thy}}: 
    G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic K from
  Bjordal's definition B:D. A4 can be proved in logic T
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  K4. Conversely, the proof that B:D follows from D1, A2, A3 and
  A4 is possible already in logic K. Hence, Bjordal's lemma
  B:L1 holds in logic S4. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual).

  \BWP{@Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?}
\paragraph{Varying domain semantics}
%\footnote{Cf. Isabelle/HOL source files \url{Bjordal/Bjordal_A/B/C_varying_domain.thy}}
: The results are
  the same as above.

>>>>

\usepackage{xspace}
\usepackage{url}
\usepackage{commands}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{modallogics}

@@

%\section{The Modal Collapse}
\noindent
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail what is called the \emph{modal collapse}
\cite{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument and stimulated attempts to remedy the problem. Hájek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996} proposed the use
of cautious instead of full comprehension principles, and Fitting \cite{fitting02:_types_tableaus_god} suggested that greater care is necessary in the semantics of higher-order quantifiers in the presence of modalities. Others, such as
Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, proposed slight emendations of
G\"odel's axioms and definitions. They require neither comprehension restrictions nor more complex semantics. Therefore, they are technically simpler to analyze with computer support. We have formalized them using the proof
assistant Isabelle/HOL \cite{Isabelle} together with the automated higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof}, and Nitpick \cite{Nitpick}.  
Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced in previous
work \cite{C40,J30,J23}. We explored the effect of different domain conditions
 on the provability of lemmas, theorems and even axioms. This was motivated by a controversy between Hájek and Anderson regarding the redundancy of some axioms in Anderson's emendation. In \emph{constant domain semantics}, the individual domains are the same in all possible worlds. In \emph{varying domain semantics}, the domains may vary from world to world. This variation is technically encoded with the help of an existence relation expressing which individuals actually exist in each world. Quantifiers are then uniformly formalized as \emph{actualist quantifiers} (i.e. guarded by the existence relation). Our main results are summarized in the following sections.

@@

For both \textbf{constant domain semantics} and \textbf{varying domain semantics}, the following results hold: 
%
G\"odel's theorems T1, C and T3' can be quickly automated (in logics \K and \KB, respectively); \BWP{3->2 respectively}
%
G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic \KFourB and the latter one already in \K).
%
Moreover, a trivial countermodel (consisting of two worlds and 
two individuals) to MC generated by Nitpick (for all mentioned logics); this model also shows the consistency of the axioms.

ToDo: tell the story of the controversy here, and then link with the mixed variant.

@@

For both \textbf{constant domain semantics} and \textbf{varying domain semantics}, the following results hold:

    G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic \K from
  Bjordal's definition B:D. A4 can be proved in logic \KT
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  \KFour. Conversely, the proof that B:D follows from D1, A2, A3 and
  A4 is possible already in logic \K. Hence, Bjordal's lemma
  B:L1 holds in logic \SFour. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic \KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual).

  \BWP{@Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?}

\BWP{ToDo: discuss the fact that Bjordal's proof is more resistant to MC than Anderson's}
